---
title: "customdatapack Vignette"
author: "Faiber Alonso falonso010@ikasle.ehu.eus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package ***customdatapack*** handles a custom class called *CustomData* that contains a data frame in the slot named *data*. *CustomData* is highly versatile, as we can initialize an object from a vector, matrix, factor, or data frame, and our class will internally convert it to a data frame. We have implemented several utilities seen as exercises in the course *"Mathematical and Statistical Software"* at the *University of the Basque Country UPV/EHU*.

To begin, the object is constructed with the constructor `customdata()`. When we print an object of *CustomData*, it will indicate that it belongs to this class and will also display the number of rows and columns of its data frame along with a preview of its first few rows, as shown below:

```{r}
library(customdatapack)

df <- data.frame(columna1 = 1:3, columna2 = letters[1:3])
obj_df <- customdata(df)
print(obj_df)

vec <- c(10, 20, 30)
obj_vec <- customdata(vec)
print(obj_vec)

fac <- factor(c("A", "B", "A"))
obj_fac <- customdata(fac)
print(obj_fac)

mat <- matrix(1:4, nrow = 2)
obj_mat <- customdata(mat)
print(obj_mat)
```

We can also use the methods `colnames()` and `column_classes()` to obtain the column names and the data class in each column:

```{r}
colnames(obj_df)
column_classes(obj_df)

colnames(obj_fac)
column_classes(obj_fac)
```

The methods `normalize()` and `standardize()` receive a *CustomData* object and return a *CustomData* object with the numeric columns normalized (from 0 to 1) or standardized (with mean 0 and standard deviation 1). Numeric columns with a single value and columns that are not numeric remain unchanged:

```{r}
df <- data.frame(a = 1:5, b = rep(5, 5), c = 10:14, d = factor(c('a','b','a','b','b')))
obj <- customdata(df)
print(obj)

normalized_obj <- normalize(obj)
print(normalized_obj)
 
standardized_obj <- standardize(obj)
print(standardized_obj)
```

The function `applyDiscretizeEF(custom_data, num.bins)` and the function `applyDiscretizeEW(custom_data, num.bins)` are two functions that receive a *CustomData* object and discretize its numeric columns into `num.bins` categories. Non-numeric columns remain unchanged. The output is a *CustomData* object with the discretized columns converted to a factor class indicating the bin to which they belong. `applyDiscretizeEF` discretizes so that each bin has approximately the same number of data points ($\pm$ 1). `applyDiscretizeEW` discretizes so that each bin has equal length. The extremes are always replaced by ($\pm$ Inf) so that these cut points can be used to discretize any other numeric vector.

```{r}
df <- data.frame(a = c(1,2,3,3,4,4,4,6,8,10), b = 1:10, c = letters[1:10])
obj <- customdata(df)

discretized_objEF <- applyDiscretizeEF(obj, num.bins = 3)
print(discretized_objEF)

discretized_objEW <- applyDiscretizeEW(obj, num.bins = 3)
print(discretized_objEW)
```

`applyEntropy()` takes a *CustomData* object and returns a *CustomData* object with entropy and normalized entropy calculated only for factor columns. Additionally, we can use `plot_entropy` on the same object to visualize the normalized entropy of factor columns in a bar chart:

```{r, fig.width=10, fig.height=6, out.width="100%"}
df <- data.frame(a = factor(c(rep('a', 95), rep('b', 5))), 
                  b = factor(c(rep('x', 50), rep('y', 50))),
                  c = factor(c(rep('k', 70), rep('l', 30))),
                  d = 1:100)
obj <- customdata(df)

entropy_values <- applyEntropy(obj)
print(entropy_values)

plot_entropy(obj)
```

We can see that the column `d`, not being a factor class, is ignored by both functions, generating a warning but not an error.

The `rocAuc` function calculates the ROC curve and AUC from a given *data.frame* or *CustomData* object containing numeric and binary label columns. The numeric label column can represent probabilities from a classifier model or real numbers to evaluate how well the data can be separated given a cut point. The binary label can be either logical values (`TRUE` or `FALSE`) or numeric values (`1` or `0`). Additionally, we can plot the ROC curve using the `plot_roc` function.


```{r, fig.width=10, fig.height=6, out.width="100%"}
set.seed(113)  # For reproducibility
df <- data.frame(
  prob = c(
    runif(50, 0.4, 1),   # High probabilities for positive class
    runif(50, 0, 0.6)    # Low probabilities for negative class
  ),
  label = c(
    rep(1, 50),  # Positive class labels
    rep(0, 50)   # Negative class labels
  )
)

obj <- customdata(df)

result <- rocAuc(obj, "prob", "label")  # Calculate ROC points and AUC

plot_roc(obj, "prob", "label") # Plot the ROC curve
```

Finally, `plotCorrMatrix` is a function that plots the correlation matrix of the numeric variables from a *data.frame* or a *CustomData* object. For this, it is strictly necessary to have the **ggplot2** and **reshape2** packages installed:

```{r, fig.width=10, fig.height=6, out.width="100%"}
set.seed(113)  # For reproducibility
df <- data.frame(a = rnorm(10), b = rnorm(10), c = rnorm(10), d = as.factor(sample(letters, 10, replace = TRUE)))
custom_data <- customdata(df)
print(df)
gg <- plotCorrMatrix(custom_data)
gg
```

We can see that columns which are not numeric (in this case, the column 'd') are not considered for the plot. The generated plot is a `ggplot` object, so it can be modified:

```{r, fig.width=10, fig.height=6, out.width="100%"}
gg + ggplot2::scale_fill_viridis_c()

```

